Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study

### 2020-07-28 pharmaco.Rmd
updated pharmacoepi manuscript

geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ./covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
csl: ../csl/styles/the-new-england-journal-of-medicine.csl
urlcolor: red

# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf"); cat(system("./wc.manuscript.sh"), "\n")

# abstract limit 250 words

@@

On behalf of Public Health Scotland COVID-19 Health Protection Study Group

@@

stepwise.drop.subparas[, 1] <- gsub("subpara\\.", "", stepwise.drop.subparas[, 1])

firstrow.neg <- which(stepwise.drop.subparas[, 3] < 0)[1]
rows.print <- firstrow.neg:nrow(stepwise.drop.subparas)

@@

_Background_ -- Dispensing of at least one prescription in the past year is a risk factor for severe COVID-19 in Scotland, even in those without diagnoses of conditions designated as conferring increased risk.

_Methods_ -- In the REACT-SCOT case-control study, ten controls were matched for sex, age and general practice to each of the `r nrow(cc.severe[CASE==1])` cases of severe COVID-19 in Scotland since the start of the epidemic.  Cases and controls were linked to records of hospital discharges since June 2015 and dispensed prescriptions issued in primary care during the last 240 days.  Associations of prescribed drugs with severe COVID-19 were examined initially for all drug classes, and subsequently for a list of 16 drug classes postulated at the start of the epidemic as "medications compromising COVID": these included proton pump inhibitors, antipsychotics, and opioids.

*Results* -- In those without a diagnosed risk condition, severe COVID-19 was strongly associated with polypharmacy; the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus none was `r table.numdrugsgr.listed.any[">12", 4]`.  This association was specifically with non-cardiovascular drugs.  Except for non-steroidal anti-inflammatory drugs, all 16 pre-specified drug classes were associated with increased risk of severe COVID-19.  The largest effect was for antipsychotic agents: rate ratio `r table.laporte["402010.Antipsychotic drugs", 3]`. Drugs used in the treatment of nausea that are pharmacologically similar to antipsychotic agents were also associated with severe disease.  Other drug classes with large effects included proton pump inhibitors (PPI), opioids and gabapentinoids.  For both PPI and opioids, there were dose-response relationships to severe COVID-19: rate ratios in those aged under 75 years were `r table.dosegr.protonpump["0-74 years: 2 or more", 4]` for $\ge 2$ DDDs/day of PPI and `r table.dosegr.opioid["0-74 years: >50", 4]` for $\ge 50$ MME opioids. The association was stronger with recent than with non-recent exposure.

*Conclusions* -- Severe COVID-19 is associated with polypharmacy and with drugs that cause sedation, respiratory depression or dyskinesia, have anticholinergic effects or affect the gastrointestinal system.  These associations are not easily explained by co-morbidity.  Although the evidence for causality is not conclusive, these results reinforce existing guidance on reducing overprescribing of these drug classes and limiting inappropriate polypharmacy.

@@

# Background
In the initial analysis of REACT-SCOT, a matched case control study of severe COVID-19 in Scotland, we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].  The univariate rate ratios associated with at least one prescription varied from 3.8 in those aged under 60 years to 2.3 in those aged 75 years and over.  This association persisted after adjusting for care home residence, hospital admission in the last five years, and diagnoses of conditions designated by public health agencies as conferring vulnerability to COVID-19 (hereafter listed conditions).  The objective of this study was to investigate this association.

# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].  In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.

For each case, the Community Health Index database was used to select up to ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.  For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.  For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.

@@

To define a prespecified hypothesis about drug classes postulated to increase the risk of severe COVID-19, we used a review by Laporte and Healy published on 2 April 2020 that listed several commonly used drug classes as associated with increased pneumonia risk and therefore of concern as "compromising COVID infections" [@laporte_midst_2020].  Excluding immunosuppressive drugs that are criteria for shielding, the Laporte-Healy list comprised proton pump inhibitors, gastrointestinal antispasmodics, H1 antihistamines, hypnotics and sedatives, antipsychotic drugs, antidepressants, drugs used in nausea and vertigo, opioid analgesics, gabapentinoids, antimuscarinic drugs used in parkinsonism, urinary antispasmodics, and non-steroidal anti-inflammatory drugs.

For proton pump inhibitors and opioids, it was possible to use equivalent doses to calculate a total dose for all drugs in the class. Defined daily doses (DDDs) of each proton pump inhibitor were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.  Morphine milligram equivalents (MMEs) for opioids were obtained from the website of the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids).  Average daily doses of proton pump inhibitors (as DDDs) and opioids (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by observation period.  The dose of opioid was calculated as the sum over opioid-containing items in subparagraph 0407010 (nonopioid and compound preparations) and 0407020 (opioid analgesics).

@@

* Adjusting for prespecified covariates and restricting the analysis to those without any  listed condition.

@@

# Results
## Relation of severe COVID-19 to polypharmacy 
We began by constructing an index of polypharmacy as the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation (from 1 June 2019 to 15 days before the date of testing positive). Table \ref{tab:numdrugs.listed.any} shows that the COVID-19 was associated with polypharmacy in those without any listed condition.  In those with a listed condition the direction of this association was reversed: those with dispensed prescriptions of 1 to 12 drug classes were at lower risk than those with no dispensed prescriptions. To study further the relationship with polypharmacy, we restricted the analysis to those without a diagnosis of a listed condition.  As prescribing of multiple drugs for primary and secondary prevention of cardiovascular disease is in accordance with evidence-based guidelines, we partitioned the number of drug classes dispensed between cardiovascular and other drugs.  In a model with two covariates -- number of cardiovascular drug classes and number of other drug classes -- the rate ratios per drug class for severe COVID-19 were respectively `r table.jointcardiovasc[1, 4]` and `r table.jointcardiovasc[2, 4]`.

## Associations with specific drug classes
Table \ref{tab:subpara.coeffs} shows univariate associations of dispensing of at least one drug in each BNF subparagraph, in those without a listed condition and filtered to show only drug classes with at least 50 exposed individuals and p < 0.001.  The drug classes associated include proton pump inhibitors, laxatives, a diverse range of drugs acting on the central nervous system, and nutritional supplements.

Most of these drugs are usually co-prescribed with other drugs and are associated with polypharmacy. To study these associations further, we abandoned the hypothesis-free approac and resorted to examining the specific drug classes listed by Laporte and Healy. Table \ref{tab:laporte} shows associations of severe COVID-19 with dispensing of each of the drug classes on the Laporte-Healy list in those aged under 75 years.  With the exception of non-steroidal anti-inflammatory drugs, all drug classes on this list were associated with increased risk in univariate analyses.  All the drug classes listed by Laporte and Healy as having anticholinergic effects -- H1 antihistamines, antidepressants, urinary antispasmodics, gastrointestinal antispasmodics, drugs for vertigo, antimuscarinic drugs used in the treatment of parkinsonism, and antiepileptic drugs -- were associated with increased risk.

In a multivariable regression, the drug classes with strongest evidence for independent association were proton pump inhibitors, antihistamines, antipsychotic drugs, and opioid analgesics.  The highest rate ratio in both univariate and multivariate analyses was that associated with antipsychotic drugs.

## Associations with antipsychotic agents and drugs used in the treatment of nausea
As others have noted [@laporte_midst_2020], the chemical structures and modes of action of drugs used in the treatment of nausea and vertigo overlap with those of drugs used as antipsychotics.  Table \ref{tab:antipsych} shows the univariate associations of severe COVID-19 with specific drugs classified in these two subparagraphs of the BNF.  Across both groups of indications, phenothiazines and other drugs that are dopamine antagonists were strongly associated with increased rates of severe disease.  The rate ratios associated with second-generation antipsychotics -- aripiprazole, olanzapine, quetiapine, and risperidone -- were as high as the rate ratios associated with phenothiazines.

** reanalyse association with drugs after excluding those likely to be receiving palliative care. **

## Associations with opioids
Table \ref{tab:dosegr} shows the relationship of severe COVID-19 to average daily opioid dose (MMEs) over the observation period.  This showed a dose-response relationship, strongest in those aged less than 75 years.  With unexposed as baseline, the univariate rate ratio in this age group was `r table.dosegr.opioid["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opioid["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile and any history of neoplasm. In those without any listed condition (all age groups combined), the unadjusted and adjusted rate ratios associated with more than 50 MME /day were `r table.dosegr.opioid.notlisted[">50", 4]` and `r table.dosegr.opioid.notlisted[">50", 6]` respectively. Table \ref{tab:timewindow} shows that the rate ratio associated with opioid dispensing only in the most recent 120-day time window was higher than that associated with dispensing only in a less recent time window (from 1 June 2019 to 121 days before the specimen date)

## Associations with proton pump inhibitors
Table \ref{tab:dosegr} shows that in those aged under 75 years the rate ratio for severe COVID-19 increases with the dose of proton pump inhibitor.  The univariate rate ratio associated with average dose of 2 or more DDDs / day was `r table.dosegr.protonpump["0-74 years: 2 or more", 4]`, reduced to `r table.dosegr.protonpump["0-74 years: 2 or more", 6]` by adjusting for care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.  This dose-response relationship was absent in those aged 75 years and over.  Among those in this age group without any listed condition, the unadjusted and adjusted rate ratios associated with more than 1.5 DDDs/day were `r table.dosegr.protonpump.notlisted.agelt75["1.5 or more", 4]` and `r table.dosegr.protonpump.notlisted.agelt75["1.5 or more", 6]` respectively. Table \ref{tab:timewindow} shows that the rate ratio associated with dispensing of a proton pump inhibitor only in the most recent 120-day time window was higher than the rate ratio associated with dispensing only in a less recent time window.

## Associations with other drug classes
Supplementary Tables \ref{tab:bnfchapter1}, \ref{tab:bnfchapter2} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 1 (gastrointestinal) 2 (cardiovascular), and 10 (musculoskeletal).  These chapters were selected as of interest because many drugs in these chapters were associated with COVID-19 or because hypotheses about possible effects of drugs in these chapters -- ACE inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.

Table \ref{tab:bnfchapter1} shows that dispensed prescriptions of most classes of drugs for the gastrointestinal system were associated with severe COVID-19, but only proton pump inhibitors, antimotility agents and laxatives were strongly and independently associated with disease in a multivariate analysis.  Proton pump inhibitors were prescribed far more commonly than any other drug class in this chapter.

@@

# Discussion 
We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation, in individuals without conditions designated as conferring high risk.  The rate ratios of 5 to 7 associated with dispensing of more than 10 drug classes are larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017].  Attempting to investigate associations with specific drugs with a hypothesis-free approach is difficult because many of the drug classes that are strongly associated with severe COVID-19, such as proton pump inhibitors, opioids and gabapentinoids, are indicators of overprescribing, recognized as such in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].

For this reason we resorted to examining a pre-specified list of drugs postulated at the start of the epidemic to increase risk of severe COVID-19, based on previously described associations with pneumonia or activity on relevant pathways, especially anticholinergic agents [@laporte_midst_2020].  We have shown that all the drugs originally listed are associated with increased risk, except for non-steroidal anti-inflammatory agents.  Distinguishing between causality and confounding as possible explanations for the association of COVID-19 with these drug classes is difficult because drug exposures are not measured accurately, and the associations are likely to be confounded by other drugs, by co-morbidities and more generally by frailty and socioeconomic deprivation.

@@

The likely mechanisms of associations of these drugs with severe COVID-19 include sedation, respiratory depression, respiratory dyskinesia and anticholinergic effects.  The evidence for these effects has been reviewed in detail elsewhere [@laporte_midst_2020] and we have little to add.  We note that as SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors and other drugs acting on the gastrointestinal tract could increase susceptibility to severe infection.  It may be relevant to investigate these associations in other countries where COVID-19 epidemics have been especially severe and overprescribing of drug classes such as proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opioids [@weiner_opioid_2017] has been reported previously.

A limitation of this study is that we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints.  Strengths of this study are that diagnoses are based on hospital discharge records coded to ICD-10 (rather than the SNOMED CT codes used in primary care databases), and that drug exposure is based on dispensed rather than issued prescriptions.

We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of drug prescribing. For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020].  The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].  Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was was markedly reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.  The numbers of cases and controls exposed to hydroxychloroquine in this dataset were too small for associations of severe COVID-19 with this drug to be estimated reliably.

## Conclusion
Severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.  As a prediction of which drug classes would be associated with increased susceptibility to severe COVID-19, the Laporte-Healy list prepared at the start of the epidemic appears to be remarkably accurate.  Most if not all of the drug classes on this list are widely overprescribed.  The consistency of associations with drugs that have similar modes of action across different groups of indication, the dose-response and time window effects support causal explanations for at least some  these associations.  We recommend that public health agencies should reinforce existing guidelines on the overprescribing of these drug classes, and more generally on the avoidance of unnecessary polypharmacy.

# Declarations
## Information governance
This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Public Health Scotland staff to link datasets.  Datasets were de-identified before analysis.  The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)). All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on [GitHub](https://github.com/pmckeigue/covid-scotland_public).  Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.

## Conflicts of interest
HC receives research support and honoraria and is a member of advisory panels or speaker’s bureaus for Sanofi Aventis, Regeneron, Novartis, Novo-Nordisk and Eli Lilly. HC receives or has recently received non-binding research support from AstraZeneca and Novo-Nordisk.  HC and PM own stock in Roche and Bayer.  SH received honoraria from Gilead.

>>>>

geometry: margin=1in
fontsize: 11pt
linenumbers: true
bibliography: ./covid.bib
date: "This version compiled `r gsub('^0', '', format(Sys.time(), '%d %B %Y'))`"
header-includes: |
  \let\origquote\quote
  \def\quote{\origquote\itshape}
  \usepackage{graphicx}
  \usepackage{longtable}
  \usepackage{booktabs}
  \usepackage{float}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{lscape}
output: 
#  pdf_document:
  rticles::nejm_article:
    includes:
      in_header: ./preamble.tex
    latex_engine: lualatex
    fig_caption: true
    keep_tex: true
    md_extensions: -autolink_bare_uris

csl: ../csl/styles/the-new-england-journal-of-medicine.csl
urlcolor: red

# rmarkdown::render("pharmaco.Rmd", output_file="pharmaco.pdf"); cat(system("./wc.manuscript.sh"), "\n"); system("make4ht -u pharmaco.tex");

# abstract limit 350 words

@@

On behalf of Public Health Scotland COVID-19 Health Protection Study Group.

@@

@@

*Objectives* -- To investigate the relation of severe COVID-19 to prior drug prescribing.

*Design* -- Matched case-control study (REACT-SCOT) based on record linkage to hospital discharges since June 2015 and dispensed prescriptions issued in primary care during the last 240 days.

*Setting* -- Scottish population.

*Main outcome measure* -- Severe COVID-19, defined by entry to critical care or fatal outcome.

*Participants* -- All `r nrow(cc.severe[CASE==1])` cases of severe COVID-19 in Scotland since the start of the epidemic, with `r nrow(cc.severe[CASE==0])` controls matched for age, sex and primary care practice.

*Results* -- Severe COVID-19 was strongly associated with the number of non-cardiovascular drug classes dispensed.  This association was strongest in those not resident in care homes, in whom the rate ratio (95% CI) associated with dispensing of 12 or more drug classes versus none was `r table.notcv.numdrugsgr.care.home[">12", 4]`, and was not accounted for by treatment of conditions designated as conferring increased risk.  Of 17 drug classes postulated at the start of the epidemic to be "medications compromising COVID", all were associated with increased risk of severe COVID-19.  The largest effect was for antipsychotic agents: rate ratio `r table.laporte["402010.Antipsychotic drugs", 3]`.  Other drug classes with large effects included proton pump inhibitors (rate rato `r table.dosegr.protonpump["0-74 years: 2 or more", 4]` for >= 2 defined daily doses/day), opioids (`r table.dosegr.opioid["0-74 years: >50", 4]` for >= 50 mg morphine equivalent/day) and gabapentinoids.  These associations persisted after adjusting for covariates, and were stronger with recent than with non-recent exposure.

*Conclusions* -- Severe COVID-19 is associated with polypharmacy and with drugs that cause sedation, respiratory depression or dyskinesia, have anticholinergic effects or affect the gastrointestinal system.  These associations are not easily explained by co-morbidity.  Although the evidence for causality is not conclusive, these results support existing guidance on reducing overprescribing of these drug classes and limiting inappropriate polypharmacy as a potential means of reducing COVID-19 risk.

*Registration* -- ENCEPP number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)

@@

***
*What is already known on this topic* -- Two previous studies have examined the relationship of severe COVID-19 to drugs for the cardiovascular system.  This is the first systematic study of the relationship of severe COVID-19 to prior drug prescribing.

*What this study adds* -- Severe COVID-19 is associated with polypharmacy and with drugs that cause sedation, respiratory depression or dyskinesia, have anticholinergic effects or affect the gastrointestinal system. These associations are not easily explained by co-morbidity.  These results support earlier warnings that these drug classes that these drugs might increase susceptibility to COVID-19, and reinforce existing guidance on reducing overprescribing of these drug classes.

***

# Background
In the initial analysis of REACT-SCOT, a matched case control study of  risk factors for severe COVID-19 in Scotland, we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].  The univariate rate ratios associated with at least one prescription varied from 3.8 in those aged under 60 years to 2.3 in those aged 75 years and over.  This association persisted after adjusting for care home residence, hospital admission in the last five years, and diagnoses of conditions designated by public health agencies as conferring vulnerability to COVID-19 (hereafter listed conditions).  The objective of this study was to investigate this association.

# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].  All record linkage studies using NHS data in Scotland, including this one, are governed by the Public Benefit and Privacy Panel for Health and Social Care, which includes patient and public representatives.  All individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.

For each case, the Community Health Index database was used to select up to ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.  For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.  For this analysis based on ascertainment of positive test results up to `r gsub("^0", "", format(max(cc.severe$SPECIMENDATE, na.rm=TRUE), "%d %B %Y"))`, entry to critical care up to `r gsub("^0", "", format(maxdate.sicsag, "%d %B %Y"))` and deaths up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[CASE==1])` cases and `r nrow(cc.severe[CASE==0])` controls.

@@

We began by testing for association of severe COVID-19 with the number of drug classes (BNF subparagraph codes) for which at least one prescription had been dispensed during the period of observation. In accordance with earlier suggestions that prescribing of multiple cardiovascular drugs, which is supported by evidence-based guidelines, should be considered separately from putatively inappropriate polypharmacy [@appleton_cardiovascular_2014], we partitioned the number of drug classes dispensed into cardiovascular and other drugs.  To define a prespecified hypothesis about drug classes postulated to increase the risk of severe COVID-19, we used a review by Laporte and Healy published on 2 April 2020 that listed 17 widely prescribed drug classes associated with increased pneumonia risk and therefore of concern as "medications compromising COVID" [@laporte_midst_2020].  Excluding immunosuppressive drugs that are criteria for shielding, the Laporte-Healy list comprised proton pump inhibitors, gastrointestinal antispasmodics, H1 antihistamines, hypnotics and sedatives, antipsychotic drugs, antidepressants, drugs used in nausea and vertigo, opioid analgesics, gabapentinoids, anti-epileptic drugs, antimuscarinic drugs used in parkinsonism, urinary antispasmodics, and non-steroidal anti-inflammatory drugs.

For proton pump inhibitors and opioids, it was possible to use equivalent doses to calculate a total dose for all drugs in the class. Defined daily doses (DDDs) of each proton pump inhibitor were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.  Morphine milligram equivalents (MMEs) for opioids were obtained from the website of the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids) of the Royal College of Anaesthestists.  Average daily doses of proton pump inhibitors (as DDDs) and opioids (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by observation period.  The dose of opioid was calculated as the sum over opioid-containing items in subparagraph 0407010 (nonopioid and compound preparations) and 0407020 (opioid analgesics).

@@

* Adjusting for prespecified covariates.

@@

# Results
## Relation of severe COVID-19 to polypharmacy 
Figure \ref{fig:notcv.numdrugs} shows that the rate of of severe COVID-19 increased steeply with the number of non-cardiovascular drug classes and decreased with the number of noncardiovascular drug classes dispensed.  Table \ref{tab:notcv.numdrugs.care.home} shows that this association was restricted to those not resident in care homes.  Supplementary Table \ref{tab:notcv.numdrugs.listed.any} shows that among those not resident in care homes, the association of severed COVID-19 with number of non-cardiovascular drug classes dispensed was present both in those without any diagnosed listed condition and in those with at least one diagnosed listed condition.

## Associations with specific drug classes
In order to identify specific drug classes associated with severe COVID-19, we examined thse associations among those not resident in a care home and without any listed condition, as most people diagnosed with listed conditions such as asthma and diabetes are on medication for these conditions.  Table \ref{tab:subpara.coeffs} shows univariate associations of dispensing of at least one drug in each BNF subparagraph, filtered to show only drug classes with at least 50 exposed individuals and $p < 0.001$.  The drug classes associated with severe COVID-19 include proton pump inhibitors, laxatives, multiple classes of drugs acting on the central nervous system, nutritional supplements, and non-steroidal anti-inflammatory drugs.

Table \ref{tab:laporte} shows associations of severe COVID-19 with dispensing of each of the drug classes on the Laporte-Healy list in those not resident in a care home. All drug classes on this list were associated with increased risk in univariate analyses (though the association with NSAIDS was stronger in Table \ref{tab:subpara.coeffs} where listed conditions were excluded.  This included all the drug classes listed by Laporte and Healy as having anticholinergic effects: H1 antihistamines, antidepressants, urinary antispasmodics, gastrointestinal antispasmodics, drugs for vertigo, antimuscarinic drugs used in the treatment of parkinsonism, and antiepileptic drugs.

In a multivariable regression, the strongest independent associations were with proton pump inhibitors, antihistamines, antipsychotic drugs, and opioid analgesics.  In both univariate and multivariable analyses, the highest rate ratio was that associated with antipsychotic drugs. As others have noted [@laporte_midst_2020], the chemical structures and modes of action of drugs used in the treatment of nausea and vertigo overlap with those of antipsychotic drugs.  Supplementary Table \ref{tab:antipsych} shows the univariate associations of severe COVID-19 with specific drugs classified in these two subparagraphs of the BNF.  Across both groups of indications, phenothiazines and other drugs that are dopamine antagonists were strongly associated with increased rates of severe disease.  Rate ratios were elevated both for phenothiazines and for second-generation antipsychotics: aripiprazole, olanzapine, risperidone and amrisulpride.

## Time window analyses
Supplementary Table \ref{tab:timewindow} shows for each drug class on the Laporte-Healy list the rate ratio associated with dispensing in the most recent 120-day time window, with dispensing only in the previous time window as reference category.  Because most users of these drugs had dispensed prescriptions in both time windows, these analyses are based on relatively small numbers: the drug classes shown are restricted to those with at least 500 cases and controls exposed only in the earlier time window .  For most of these drug classes shown, the rate ratio associated with recent exposure only is above 1, but only for opioid analgesics, and to some extent proton pump inhibitors and anxiolytics, does this association reach conventional levels of statistical significance.

## Dose-response analyses
Supplementary Table \ref{tab:dosegr} shows the relationship of severe COVID-19 to average daily doses of opioids (as MME) and proton pump inhibitors (as DDDs) over the 240-day observation period.  For both these drug classes there were dose-response relationships, and for proton pump inhibitors this relationship was strongest in those aged less than 75 years.  With unexposed as baseline, the univariate rate ratio associated with opioid use in this age group was `r table.dosegr.opioid["0-74 years: >50", 4]` in those with average daily dose of more than 50 mg morphine equivalent (MME), reduced to `r table.dosegr.opioid["0-74 years: >50", 6]` on adjusting for care home residence, SIMD quintile and any history of neoplasm. 
For proton pump inhibitors, the univariate rate ratio associated with average dose of 2 or more DDDs / day was `r table.dosegr.protonpump["0-74 years: 2 or more", 4]`, reduced to `r table.dosegr.protonpump["0-74 years: 2 or more", 6]` by adjusting for care home residence, SIMD quintile, any diagnosis of ICD-10 codes K20-K31 (diseases of esophagus, stomach and duodenum), non-steroidal anti-inflammatory drugs, anti-platelet agents and anticoagulants.

## Associations with other drug classes
Supplementary Tables \ref{tab:bnfchapter2} and \ref{tab:bnfchapter10} show associations with drug classes in BNF chapters 2 (cardiovascular) and 10 (musculoskeletal).  These chapters were selected as of interest because because specific hypotheses about possible effects of drugs in these chapters -- ACE inhibitors [@mancia_reninangiotensinaldosterone_2020], anticoagulants [@connors_covid-19_2020], and hydroxychloroquine [@alexander_covid-19_2020]-- have been proposed or discussed.

@@

# Discussion 
<!---
• Statement of principal findings
• Strengths and weaknesses of the study
• Strengths and weaknesses in relation to other studies, discussing important differences in results
• Meaning of the study: possible explanations and implications for clinicians and policymakers
• Unanswered questions and future research
--->
We have shown that severe COVID-19 is associated with polypharmacy, defined by the number of drugs classes dispensed during the period of observation, in individuals without conditions designated as conferring high risk.  The rate ratios of 5 to 7 associated with dispensing of more than 10 drug classes are larger than the ratio of about 2 for all-cause mortality associated with this level of polypharmacy in a systematic review [@leelakanok_association_2017].  Attempting to investigate associations with specific drugs with a hypothesis-free approach is difficult because many of the drug classes that are strongly associated with severe COVID-19, such as proton pump inhibitors, opioids and gabapentinoids, are indicators of overprescribing, recognized as such in the Scottish National Therapeutic Index of prescribing quality [@macbride-stewart_national_2019].

To narrow the hypothesis space we tested a pre-specified list of drugs postulated at the start of the epidemic to increase risk of severe COVID-19, based on previously described associations with pneumonia or activity on relevant pathways, especially anticholinergic agents [@laporte_midst_2020].  We have shown that all the drugs originally listed are associated with increased risk. Distinguishing between causality and confounding as possible explanations for the association of COVID-19 with these drug classes is difficult because drug exposures are not measured accurately, and the associations are likely to be confounded by other drugs, by co-morbidities and more generally by frailty and socioeconomic deprivation.

@@

A limitation of this study is that we do not have morbidity data from primary care, which would include risk factors such as smoking and coding of presenting complaints. Another limitation is that it is not possible to capture hospital prescribing data which includes biologic agents that have immunosuppressive effects.  Strengths of this study are that diagnoses are based on hospital discharge records coded to ICD-10 (rather than the SNOMED-CT codes used in primary care databases), and that drug exposure is based on dispensed rather than issued prescriptions.

The mechanisms postulated by Laporte and Healy for drugs to increase risk of severe COVID-19 include sedation, respiratory depression, respiratory dyskinesia and anticholinergic effects  [@laporte_midst_2020].  We note that as SARS-CoV-2 is at least partly an enteric infection [@wan_enteric_2020] and the ACE2 receptor is expressed in the intestine, it is plausible that proton pump inhibitors and other drugs acting on the gastrointestinal tract could increase susceptibility to severe infection.  It may be relevant to investigate these associations in other countries where COVID-19 epidemics have been especially severe and overprescribing of drug classes such as proton pump inhibitors [@luo_changes_2018; @pasina_overuse_2020; @ramirez_overuse_2010] or opioids [@weiner_opioid_2017] has been reported previously.

We emphasize that because of the relationship of COVID-19 to polypharmacy, associations with specific drug classes cannot be studied without taking into account how those drug classes are related to the profile of drug prescribing. Of those drug classes that are not on the Laporte-Healy list, there are strong univariate associations of severe COVID-10 with dispensing of antibiotics, laxatives and nutritional supplements. The associations with nutritional supplements is likely to be confounded by overprescribing. For ACE inhibitors and angiotensin-II receptor blockers, our results are consistent with other studies that have found no increased risk associated with these drugs [@mancia_reninangiotensinaldosterone_2020; @de_abajo_use_2020] and indeed suggest some protective effect may be possible.  To explore this more fully will require access to other datasets where measurements of blood pressure and other covariates are available. The relation of anticoagulant use to COVID-19 is of interest because coagulopathy is a feature of severe disease [@connors_covid-19_2020].  Although in this study anticoagulants were associated with increased risk in univariate analysis, this association was reduced by adjusting for covariates including diagnosed heart disease and co-prescribing of proton pump inhibitors.  Associations with disease-modifying antirheumatic drugs such as hydroxychloroquine are likely to be confounded by hospital-based prescribing of biologic anti-rheumatic drugs, which is not captured by record linkage in Scotland.

## Conclusion
Severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.  As a prediction of which drug classes would be associated with increased susceptibility to severe COVID-19, the Laporte-Healy list prepared at the start of the epidemic appears to be remarkably accurate.  Many of the drug classes on this list are recognized indicators of overprescribing.  The consistency of associations with drugs that have similar modes of action across different groups of indication, the dose-response and time window effects support causal explanations for at least some these associations.  We recommend that public health agencies should reinforce existing guidelines on avoidance of overprescribing of these drug classes and more generally on inappropriate polypharmacy.

# Declarations
## Ethics approval and information governance
This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Public Health Scotland staff to link datasets.  Datasets were de-identified before analysis.  Dissemination of results will be managed by Public Health Scotland.

## Data sharing
Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.

## Role of the funding source
This study was undertaken without any external funding.
